quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-06-04 paper-quollr.Rmd
Select the dataset to evaluate from cardinalR

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)
library(Rtsne)
library(readr)
library(tools)

@@

)
```

@@

## Constructing the 2-D model

@@

@@

<!--with prism data-->

>>>>

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)
library(Rtsne)
library(readr)
library(tools)
library(cardinalR)

@@

)

interior_annotation <- function(label, position = c(0.92, 0.92)) {
  annotation_custom(grid::textGrob(label = label,
      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
      gp = grid::gpar(cex = 1, col="grey70")))
}
```

@@

## Constructing the $2\text{-}D$ model

Constructing the $2\text{-}D$ model mainly contains (i) scaling, (ii) computing hexagon grid configurations, (iii) binning, and (iv) indicating neighbors by line segments connecting centroids.

@@

```{r}
scaled_nldr_data <- gen_scaled_data(data = s_curve_noise_umap)

scaled_nldr_data
```

@@

<!--with prism data-->
<!-- my plan is to discuss which the appropriate representation from different parameter choices from tSNE-->
<!--Why tSNE? thinking about the usability-->

```{r}
triangular_3d_data <- tri_3d(
  n = 1250, num_noise = 2, 
  min_n = -0.05, max_n = 0.05) |>
  as_tibble()

colnames(triangular_3d_data) <- paste0("x", 1:NCOL(triangular_3d_data))

langevitour::langevitour(triangular_3d_data)
```

```{r}
sc_ltr_pos <- c(0.08, 0.96)

triangular_3d_data <- triangular_3d_data |>
  mutate(ID = row_number())

tSNE_fit <- triangular_3d_data |>
  select(-ID) |>
  Rtsne(perplexity = sqrt(NROW(triangular_3d_data)))

tsne_prism <- tSNE_fit$Y |>
  as_tibble() |>
  mutate(ID = row_number())

colnames(tsne_prism) <- c("tSNE1", "tSNE2", "ID")

prism_scaled_obj <- gen_scaled_data(
  data = tsne_prism)
tsne_prism_scaled <- prism_scaled_obj$scaled_nldr

nldr_prism <- tsne_prism_scaled |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5)

nldr_prism
```

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_prism <- 2:25

lim1 <- prism_scaled_obj$lim1
lim2 <- prism_scaled_obj$lim2
r2_prism <- diff(lim2)/diff(lim1)

error_prism <- data.frame(matrix(nrow = 0, ncol = 0))

for (xbins in bin1_vec_prism) {

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2_prism)$bin2

  prism_model <- fit_highd_model(
  training_data = triangular_3d_data,
  emb_df = tsne_prism_scaled,
  bin1 = xbins,
  r2 = r2_prism,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "x"
)

  df_bin_centroids_prism <- prism_model$df_bin_centroids
  df_bin_prism <- prism_model$df_bin

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_prism,
    df_bin = df_bin_prism,
    training_data = triangular_3d_data,
    newdata = NULL,
    type_NLDR = "tSNE",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_prism))

  error_prism <- bind_rows(error_prism, error_df)

}

mse_prism_b <- ggplot(error_prism, 
                     aes(x = b, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 420, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  theme(axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90))

mse_prism_b
```

```{r}
## Compute hexbin parameters
num_bins_x_prism <- 20
lim1 <- prism_scaled_obj$lim1
lim2 <- prism_scaled_obj$lim2
r2_prism <- diff(lim2)/diff(lim1)

prism_model <- fit_highd_model(
  training_data = triangular_3d_data,
  emb_df = tsne_prism_scaled,
  bin1 = num_bins_x_prism,
  r2 = r2_prism,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "x"
)

df_bin_centroids_prism <- prism_model$df_bin_centroids
df_bin_prism <- prism_model$df_bin

## Triangulate bin centroids
tr1_object_prism <- tri_bin_centroids(
  df_bin_centroids_prism, x = "c_x", y = "c_y")
tr_from_to_df_prism <- gen_edges(
  tri_object = tr1_object_prism)

## Compute 2D distances
distance_prism <- cal_2d_dist(
  tr_coord_df = tr_from_to_df_prism,
  start_x = "x_from",
  start_y = "y_from",
  end_x = "x_to",
  end_y = "y_to",
  select_vars = c("from", "to", "distance"))

## To find the benchmark value
benchmark_prism <- find_lg_benchmark(
  distance_edges = distance_prism,
  distance_col = "distance")

trimesh_removed_prism <- vis_rmlg_mesh(
  distance_edges = distance_prism,
  benchmark_value = benchmark_prism,
  tr_coord_df = tr_from_to_df_prism,
  distance_col = "distance") #+
  #xlim(sc_xlims) + ylim(sc_ylims) +
  #interior_annotation("a", sc_ltr_pos)

trimesh_removed_prism
```

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_prism <- hex_binning(
  data = tsne_prism_scaled,
  bin1 = num_bins_x_prism,
  r2 = r2_prism)

tsne_data_with_hb_id <- hb_obj_prism$data_hb_id

df_all_prism <- dplyr::bind_cols(triangular_3d_data |> dplyr::select(-ID),
                                tsne_data_with_hb_id)

### Define type column
df <- df_all_prism |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset

df_b <- df_bin_prism |>
  dplyr::filter(hb_id %in% df_bin_centroids_prism$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_prism$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)

df_exe <- dplyr::bind_rows(df_b, df)

## Set the maximum difference as the criteria
distance_df_small_edges <- distance_prism |>
  dplyr::filter(distance < benchmark_prism)
## Since erase brushing is considerd.

langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(0.8, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```
